---
title: "Statistical distribution functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with `dist_fns` Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: TRUE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

```{r setup}
library(tidyabc)
library(dplyr)
library(ggplot2)

ggplot2::set_theme(theme_minimal())

```

## Introduction

There are parts to this vignette. A description of extra distribution functions
available in `tidyabc` and explanation of the `dist_fns` S3 class which collects
statistical distribution functions for use as a single entity.

## Reparameterised statistical distributions

The `tidyabc` package provides several useful probability distribution functions
beyond the standard R families available through `stats`. These are particularly
handy for modelling scenarios common in ABC workflows, offering convenient
parametrizations (e.g., by mean and standard deviation) or specific support
ranges.

### Core Functions

Like standard R distributions (e.g., `stats::dnorm`, `stats::pnorm`), these
custom families provide four main types of functions:

- **`dXXX(x, ...)`**: Density (PDF) or Probability Mass (PMF) function.
- **`pXXX(q, ...)`**: Cumulative Distribution Function (CDF).
- **`qXXX(p, ...)`**: Quantile function (inverse CDF).
- **`rXXX(n, ...)`**: Random number generation function.

### Selected Examples

Here are examples of a few key families available in `tidyabc`:

**1. Logit-Normal (`logitnorm` / `logitnorm2`)**

A distribution supported on (0, 1), useful for modelling probabilities or
proportions. The `logitnorm2` family parametrizes via the median (`prob.0.5`)
and a dispersion parameter (`kappa`).

```{r}
# Density at x=0.7 with median 0.6 and low dispersion
dlogitnorm2(0.7, prob.0.5 = 0.6, kappa = 0.2)

# CDF value at q=0.5
plogitnorm2(0.5, prob.0.5 = 0.6, kappa = 0.2)

# Generate 5 random samples
rlogitnorm2(5, prob.0.5 = 0.6, kappa = 0.2)
```

**2. Beta (`beta2`)**

Another distribution supported on (0, 1), re-parametrized for mean (`prob`) and
a coefficient of variation (`kappa`), useful for modelling uncertainty around a
central probability.

```{r}
# Density at x=0.3 with mean 0.3 and variability 0.1
dbeta2(0.3, prob = 0.3, kappa = 0.1)

# Quantile for the 90th percentile
qbeta2(0.9, prob = 0.3, kappa = 0.1)

# Generate 5 random samples
rbeta2(5, prob = 0.3, kappa = 0.1)
```

**3. Gamma (`gamma2`, `cgamma`)**

Distributions supported on (0, ∞). `gamma2` is parametrized by mean and standard
deviation. `cgamma` is a "convex" gamma (mean > standard deviation), useful for
positive variables with a defined mode.

```{r}
# Density at x=6 for a gamma with mean=5, sd=2
dgamma2(6, mean = 5, sd = 2)

# Quantile for the 75th percentile for convex gamma
qcgamma(0.75, mean = 3, kappa = 0.4) # kappa is coef. of variation here

# Generate 5 random samples from gamma2
rgamma2(5, mean = 5, sd = 2)
```

**4. Wedge (`wedge`)**

A distribution supported on [0, 1] with a linear probability density function.
The `a` parameter controls skewness (a=0 is uniform, a>0 is right-skewed, a<0 is
left-skewed).

```{r}
# Density at x=0.8 with right skew (a=1)
dwedge(0.8, a = 1)

# Quantile for the 50th percentile (median) with left skew (a=-1)
qwedge(0.5, a = -1)

# Generate 5 random samples
rwedge(5, a = 1)
```

### List of Additional Families

- `logitnorm`, `logitnorm2`
- `wedge`
- `beta2`
- `lnorm2` (Log-Normal)
- `gamma2`, `cgamma` (Convex Gamma)
- `nbinom2` (Negative Binomial)
- `null` (Always returns `NA` - via `rnull`, `pnull`, etc.)

These functions can be used directly in R for calculations, simulations, or
defining priors before wrapping them into `dist_fns` objects.

### Specialized Random Number Generators

`tidyabc` also provides specific functions designed primarily for random
sampling, which are particularly useful for forward simulation within ABC
workflows.

**1. Bernoulli (`rbern`)**

Generates logical values (`TRUE`/`FALSE`) based on a probability of success.

```{r}
# Simulate 10 Bernoulli trials with prob=0.3
rbern(10, prob = 0.3)
```

**2. Categorical (`rcategorical`)**

Generates random samples from a multinomial-like distribution over a set of
named or numbered categories, based on provided probabilities.

```{r}
# Define probabilities for categories "A", "B", "C"
probs <- c("A" = 0.1, "B" = 0.3, "C" = 0.6)

# Sample 5 categories according to the probabilities
rcategorical(5, prob = probs)

# Sample 5 categories and return them as a factor
rcategorical(5, prob = probs, factor = TRUE)
```

**3. Exponential Growth Process Times (`rexpgrowth`, `rexpgrowthI0`)**

These functions sample event times from an exponentially growing (or decaying)
process. This is useful for simulating arrival times, infection times, or other
phenomena where the underlying rate changes exponentially over time.

- `rexpgrowth(n, r, t_end, t_start)`: Generates `n` event times assuming an 
exponential growth rate `r` over the interval `[t_start, t_end]`.
- `rexpgrowthI0(I0, r, t_end, t_start)`: Generates event times aiming for an 
*expected* initial number of events `I0` in the first unit of time, 
given the growth rate `r`.

```{r}
# Example: Simulate 100 event times from a process growing at rate 0.1
# over the interval [0, 40]
event_times_growth <- rexpgrowth(100, r = 0.1, t_end = 40, t_start = 0)
hist(event_times_growth, breaks = 20, main = "Event Times (Growth r=0.1)")

# Example: Simulate event times aiming for an initial rate of 10 events per day
# over 20 days with a growth rate of 0.05
event_times_I0 <- rexpgrowthI0(I0 = 10, r = 0.05, t_end = 20, t_start = 0)
hist(event_times_I0, breaks = 20, main = "Event Times (I0=10, r=0.05)")
```

These functions provide convenient ways to model specific random processes often
encountered in simulation models used with ABC.

---

# Distribution family S3 class:

The `dist_fns` S3 class in `tidyabc` provides a powerful and unified interface
for working with probability distributions. It encapsulates the core statistical
functions (CDF `p`, quantile `q`, density `d`, random `r`) into a single object,
enabling flexible creation, manipulation, and analysis. This vignette
demonstrates how to create `dist_fns` objects from standard families, manipulate
them within data frames using `tidyverse` principles, fit empirical
distributions from data or quantiles, and combine them into mixtures.

## Creating `dist_fns` Objects

### From Standard Families

The most straightforward way to create a `dist_fns` object is from standard R distribution families using `as.dist_fns()`.

```{r}
# Create a single normal distribution object
norm_dist <- as.dist_fns("norm", mean = 4, sd = 3)
print(norm_dist)

# Access its functions directly
norm_dist$p(5) # CDF at x=5
norm_dist$q(0.95) # 95th percentile
norm_dist$r(5) # Generate 5 random samples
```

### Creating Multiple Distributions with `pmap_dist_fns`

You can create a *list* of `dist_fns` objects, each with different
parameters. This is efficiently done using `pmap_dist_fns()` within a
`dplyr::mutate()` call, mapping parameter values from columns of a data frame.

```{r}
# Define a data frame with parameters for different families
param_df <- tibble(
  id = 1:3,
  family = c("norm", "gamma2", "lnorm2"),
  mean = c(0, 2, 5),
  sd = c(1, 1.5, 2)
)

# Use pmap_dist_fns to create a list column of dist_fns objects
dist_df <- param_df %>%
  mutate(
    dist_obj = pmap_dist_fns(
      .,
      function(family, mean, sd, ...) as.dist_fns(family, mean, sd)
    )
  )

# N.b. annoyingly RStudio does not use the default pillar printer and the
# details do not show up when rendering a tibble in RStudio
print(dist_df)

```

This pattern is intended for scenarios like storing prior distributions or
posterior approximations for multiple parameters within a single data frame.

Plotting distributions together is managed with the `plot` function which can
be modified with `ggplot` commands.

```{r}

plot(dist_df$dist_obj)

```

---

## Manipulating `dist_fns` Objects in Data Frames

Once `dist_fns` objects are stored in a data frame column, you can manipulate them using `map_dist_fns()` or standard `purrr`/`dplyr` functions.

### Applying Transformations: Truncation Example

You can apply transformations like `truncate()` to `dist_fns` objects stored in a column.

```{r}
# Truncate each distribution from the previous example to be positive
dist_df_truncated <- dist_df %>%
  mutate(
    dist_truncated = map_dist_fns(dist_obj, ~ truncate(.x, x_left = 0, x_right = 5))
  )

# Plot the truncated versions of the distributions
plot(dist_df_truncated$dist_truncated)
```

---

## Combining `dist_fns`: Mixture Distributions

A powerful feature is creating mixture distributions from a list of `dist_fns` objects, for example, using `dplyr::summarise()`.

```{r}
# Create a list of different distribution objects (e.g., representing different scenarios/models)

mix_df = dist_df %>% 
  # Assign weights to each component
  mutate(weights = c(0.4, 0.4, 0.2)) %>%
  summarise(
    mix = c(mixture(dist_obj, weights = weights, name="weighted"))
  )

mixture_dist = mix_df$mix

# Plot the mixture
plot(mixture_dist)

# Evaluate the mixture CDF
mixture_dist$p(c(0, 1, 2, 3, 4))
```

---

## Fitting Empirical Distributions

`tidyabc` provides robust tools for fitting empirical distributions from data or quantiles.

### From Quantiles using `empirical_cdf`

If you have quantile values (e.g., from another distribution or simulation
output), you can fit an empirical CDF. This takes a link parameter which lets
you easily define support for the empirical CDF. The resulting fit will
impute the tails of the distribution to prevent truncation. 

```{r}
# Example: Fit an empirical distribution to quantiles of a known distribution
target_dist <- as.dist_fns("gamma", shape = 2, rate = 1)
# Define probability points
p_points <- c(0.025, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.975) 
# Get corresponding quantiles
q_values <- target_dist$q(p_points)

# Fit an empirical distribution using these quantiles
empirical_from_quantiles <- empirical_cdf(x = q_values, p = p_points, link = "log", name = "Empirical from Quantiles")

# Compare the original and fitted 
plot(c(target_dist,empirical_from_quantiles))

```

#### Ensemble of Quantile-Based Distributions as a Mixture

You can create an ensemble of such quantile-based fits and combine them into a mixture.

```{r}

# From previous example we create a data frame with quantiles in it
quantile_df = dist_df %>% 
  mutate(
    quantiles = purrr::map(dist_obj, ~ .x$q(p_points))
  ) %>% 
  select(-dist_obj)

# now if this was our input we could fit empirical distributions:
# N.B. We could do this with a grouped data frame.
ensemble_df = quantile_df %>%
  mutate(
    emp_cdf = map_dist_fns(quantiles, function(q) empirical(x=q,p = p_points))
  ) %>%
  summarise(
    ensemble = c(mixture(emp_cdf,name = "unweighted ensemble"))
  )


plot(c(ensemble_df$ensemble, mixture_dist))
```

### From Samples using `empirical_data`

For fitting directly from sample data (e.g., MCMC output, simulation results),
use `empirical()` or `empirical_data()`.

```{r}
# Generate sample data from a distribution
set.seed(123)
sample_data <- rgamma(1000, shape = 2, rate = 1)

# Fit an empirical distribution to the samples
empirical_from_samples <- empirical(
  x = sample_data, name = "Empirical from Samples", bw=0.25)

# Compare the sample histogram with the fitted density
tibble(sample = sample_data) %>%
  ggplot(aes(x = sample)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.5, fill = "lightblue", color = "black") +
  stat_function(fun = empirical_from_samples$d, color = "red", linewidth = 1) +
  labs(title = "Sample Histogram vs. Fitted Empirical Density", x = "x", y = "Density")
```

#### Weighted Sample Fitting

`empirical()` also handles *weighted* samples, which is crucial in methods like
ABC where particles have importance weights.

```{r}
# Simulate weighted samples (e.g., importance sampling output)

set.seed(456)
n_samples <- 500

raw_samples <- c(
  rnorm(n_samples / 2, mean = 1, sd = 0.5), 
  rnorm(n_samples / 2, mean = 3, sd = 0.7)
)

# Simulate some weights (e.g., importance weights)
# these are correlated to the distance from 3, which for example might be the 
# true value
weights <- runif(n_samples, min = 0.1, max = 2.0) * exp(-((raw_samples - 3)/2)^2)

# Normalize weights (often done automatically internally, but good practice)
weights <- weights / sum(weights)

# Fit empirical distribution from weighted samples
empirical_weighted <- empirical(
  x = raw_samples, w = weights, 
  name = "weighted", bw=0.25
)

empirical_unweighted <- empirical(
  x = raw_samples, 
  name = "unweighted", bw=0.25
)

# Plot the fitted distribution
plot(c(empirical_weighted,empirical_unweighted))
```

#### Using `wquantile` for Quantiles from Weighted Data

The approach above can give you quantiles from the data. It is however fairly
heavyweight if you know you are only going ot need a few quantiles from the
weighted data.

The `wquantile()` function provides a quicker way to estimate quantiles from
weighted data, including use of link functions, using a similar but less complex
approach to `empirical`. In this case agreement is close but not perfect

```{r}
# Calculate specific quantiles from the weighted samples using wquantile
weighted_quantiles <- wquantile(
  p = c(0.025, 0.25, 0.5, 0.75, 0.975), # Common quantiles
  x = raw_samples,
  w = weights,
  link = "log", # Use identity link for unbounded support, or specify prior support
)

print(tibble(Probability = c(0.025, 0.25, 0.5, 0.75, 0.975), Quantile = weighted_quantiles))

# Compare with quantiles from the fitted empirical distribution (which used the same weighted data)
fitted_quantiles <- empirical_weighted$q(c(0.025, 0.25, 0.5, 0.75, 0.975))

print(tibble(Probability = c(0.025, 0.25, 0.5, 0.75, 0.975), Quantile_Fitted = fitted_quantiles))
# Values should be similar.
# Choices regarding the degree of smoothing will affect the exact answer.
```

---

## Conclusion

The `dist_fns` class provides a flexible and consistent way to represent,
manipulate, and combine probability distributions in R, especially within a
`tidyverse` workflow. Its integration with data frames allows for powerful batch
operations and storage of complex distributional information. The empirical
fitting capabilities make it suitable for approximating complex distributions
derived from data or simulation results, including handling weighted samples
common in Bayesian and ABC methods.

